---
title: "Over-Representation Analysis with ClusterProfiler (modified by me)"
author: "Mohammed Khalfan (modified by Ana-Mariya Anhel Valdes)"
date: "5/15/2019"
output:
  html_document:
    df_print: paged
df_print: paged
---

This R Notebook describes the implementation of over-representation analysis using the clusterProfiler package. For more information please see the full documentation here: https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html

# Set the data folder
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "folder_data")
```

# Install and load packages
```{r, message=F, warning=F}
#install.packages("BiocManager")
#BiocManager::install("clusterProfiler")
#BiocManager::install("pathview")
#install.packages("wordcloud")
library(clusterProfiler)
library(wordcloud)
library(ggplot2) # to make the plots a little bit more pretty
```

# Annotations
I'm using *D melanogaster* data, so I install and load the annotation "org.Dm.eg.db" below. See all annotations available here: http://bioconductor.org/packages/release/BiocViews.html#___OrgDb (there are 19 presently available). 
In my case I am going to use the human species because my genes are filtered based on that
the organism is "org.Hs.eg.db" --> https://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html

```{r, message=F, warning=F}
#organism = "org.Dm.eg.db"
organism = "org.Hs.eg.db"
#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)
```


#Prepare Input

```{r}
# reading in input from deseq2
manually_labelled <- read.csv("training-test-held-out-metadata-labels-formatted_withmygeneIDs_unified.csv", header = TRUE)
final <- read.csv("finalSearch-dataset-metadata-labels-formatted_withmygeneIDs_unified.csv", header = TRUE)
final_filtered <- read.csv("finalSearch-dataset-metadata-labels-formatted_withmygeneIDs_unified_filtered.csv", header = TRUE)
```


# separate regulatros from non-regulators
```{r}
reg_manual <- as.character(manually_labelled$id_human[manually_labelled$labels == 1])
non_reg_manual <- as.character(manually_labelled$id_human[manually_labelled$labels == 0])

reg_final <- as.character(final$id_human[final$predicted_label == 1])
non_reg_final <- as.character(final$id_human[final$predicted_label == 0])
  
reg_final_filt <- as.character(final_filtered$id_human[final_filtered$predicted_label == 1])
non_reg_final_filt <- as.character(final_filtered$id_human[final_filtered$predicted_label == 0])
```

#Create enrichGO object
Params:  
  
**Ontology** Options: ["BP", "MF", "CC"]  biological processes, molecular functions and cellular components
**keyType** This is the source of the annotation (gene ids). The options vary for each annotation. In the example of *org.Dm.eg.db*, the options are:   
  
"ACCNUM"       "ALIAS"        "ENSEMBL"      "ENSEMBLPROT"  "ENSEMBLTRANS" "ENTREZID"      
"ENZYME"       "EVIDENCE"     "EVIDENCEALL"  "FLYBASE"      "FLYBASECG"    "FLYBASEPROT"   
"GENENAME"     "GO"           "GOALL"        "MAP"          "ONTOLOGY"     "ONTOLOGYALL"   
"PATH"         "PMID"         "REFSEQ"       "SYMBOL"       "UNIGENE"      "UNIPROT"  
  
Check which options are available with the `keytypes` command, for example `keytypes(org.Dm.eg.db)`. 

```{r}
keytypes(org.Hs.eg.db)
```
Very similar to drosophila to be honest
```{r}
org.Hs.eg
```
## Create the objects

https://rdrr.io/bioc/clusterProfiler/man/enrichGO.html
this is where I got to see what each argument is

We are not going to use *universe* because we want to compare with all the genes
we want to see "are these genes enriched compared to all human genes?"

We are going, for the first analysis use the default ones of *minGSSize* and *maxGSSize* but if we dont get a lot of results we can go lower to 5 in minGSSize

### manual labelled
```{r}
go_enrich_reg <- enrichGO(gene = reg_manual,
                      OrgDb = org.Hs.eg.db, # human, as we set it before
                      keyType = 'ENTREZID', # our ids are entrez
                      readable = T, # mapping id to gene name
                      ont = "BP", # We want to asses biological processes
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10,
                      pAdjustMethod = "BH") # to control FDR

go_enrich_non_reg <- enrichGO(gene = non_reg_manual,
                      OrgDb = org.Hs.eg.db, # human, as we set it before
                      keyType = 'ENTREZID', # our ids are entrez
                      readable = T, # mapping id to gene name
                      ont = "BP", # We want to asses biological processes
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10,
                      pAdjustMethod = "BH") # to control FDR
```

```{r}
ora_simplified <- simplify(go_enrich_reg, 
                           cutoff = 0.7, 
                           by = "p.adjust")
dotplot(ora_simplified, showCategory = 10)
```

```{r}
# make the plot and store it
a <- dotplot(go_enrich_reg, 
             showCategory = 10, 
             title = "GO Biological Process - Regulators - Manual Curation") +
  theme_bw() +                                    # cleaner background
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),          # GO term labels
    axis.text.x = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12)
  ) +
  #scale_colour_gradient(low = "red", high = "green") # low p = red = significant
  scale_colour_viridis_c(option = "viridis", direction = -1)+
  guides(colour = guide_colourbar())

b <- dotplot(go_enrich_non_reg, 
             showCategory = 10, 
             title = "GO Biological Process - No Regulators - Manual Curation") +
  theme_bw() +                                    # cleaner background
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),          # GO term labels
    axis.text.x = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12)
  ) +
  #scale_colour_gradient(low = "red", high = "green") # low p = red = significant
  scale_colour_viridis_c(option = "viridis", direction = -1)+
  guides(colour = guide_colourbar())
```

### final dataset
```{r}
go_enrich_reg_final <- enrichGO(gene = reg_final,
                      OrgDb = org.Hs.eg.db, # human, as we set it before
                      keyType = 'ENTREZID', # our ids are entrez
                      readable = T, # mapping id to gene name
                      ont = "BP", # We want to asses biological processes
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10,
                      pAdjustMethod = "BH") # to control FDR

go_enrich_non_reg_final <- enrichGO(gene = non_reg_final,
                      OrgDb = org.Hs.eg.db, # human, as we set it before
                      keyType = 'ENTREZID', # our ids are entrez
                      readable = T, # mapping id to gene name
                      ont = "BP", # We want to asses biological processes
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10,
                      pAdjustMethod = "BH") # to control FDR
```

```{r}
# make the plot and store it
c <- dotplot(go_enrich_reg_final, 
             showCategory = 10, 
             title = "GO Biological Process - Regulators - Final prediction") +
  theme_bw() +                                    # cleaner background
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),          # GO term labels
    axis.text.x = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12)
  ) +
  #scale_colour_gradient(low = "red", high = "green") # low p = red = significant
  scale_colour_viridis_c(option = "viridis", direction = -1)+
  guides(colour = guide_colourbar())

d <- dotplot(go_enrich_non_reg_final, 
             showCategory = 10, 
             title = "GO Biological Process - No Regulators - Final prediction") +
  theme_bw() +                                    # cleaner background
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),          # GO term labels
    axis.text.x = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12)
  ) +
  #scale_colour_gradient(low = "red", high = "green") # low p = red = significant
  scale_colour_viridis_c(option = "viridis", direction = -1)+
  guides(colour = guide_colourbar())
```

### final dataset filtered
```{r}
go_enrich_reg_final_filt <- enrichGO(gene = reg_final_filt,
                      OrgDb = org.Hs.eg.db, # human, as we set it before
                      keyType = 'ENTREZID', # our ids are entrez
                      readable = T, # mapping id to gene name
                      ont = "BP", # We want to asses biological processes
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10,
                      pAdjustMethod = "BH") # to control FDR

go_enrich_non_reg_final_filt <- enrichGO(gene = non_reg_final_filt,
                      OrgDb = org.Hs.eg.db, # human, as we set it before
                      keyType = 'ENTREZID', # our ids are entrez
                      readable = T, # mapping id to gene name
                      ont = "BP", # We want to asses biological processes
                      pvalueCutoff = 0.05, 
                      qvalueCutoff = 0.10,
                      pAdjustMethod = "BH") # to control FDR
```

```{r}
# make the plot and store it
e <- dotplot(go_enrich_reg_final_filt, 
             showCategory = 10, 
             title = "GO Biological Process - Regulators - Final prediction Filtered") +
  theme_bw() +                                    # cleaner background
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),          # GO term labels
    axis.text.x = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12)
  ) +
  #scale_colour_gradient(low = "red", high = "green") # low p = red = significant
  scale_colour_viridis_c(option = "viridis", direction = -1)+
  guides(colour = guide_colourbar())

f <- dotplot(go_enrich_non_reg_final_filt, 
             showCategory = 10, 
             title = "GO Biological Process - No Regulators - Final prediction Filtered") +
  theme_bw() +                                    # cleaner background
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 16),
    axis.text.y = element_text(size = 14),          # GO term labels
    axis.text.x = element_text(size = 14),
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 12)
  ) +
  #scale_colour_gradient(low = "red", high = "green") # low p = red = significant
  scale_colour_viridis_c(option = "viridis", direction = -1)+
  guides(colour = guide_colourbar())
```

```{r}
#library(stringr)
library(scales)

e_v2 <- dotplot(
  go_enrich_reg_final_filt, 
  showCategory = 10,
  title = NULL,
  color = "p.adjust",  # Keep statistical variable here
  label_format = 45,
  #orderBy = "p.adjust" # order the terms by p-adj instead of gene ratio
) +
  # Apply Viridis palette to BOTH color and fill aesthetics simultaneously
  scale_color_viridis_c(
    option = "viridis", 
    direction = -1,
    name = expression(italic(p)[adj]),
    #labels = label_number(accuracy = 0.01), # Rounds numbers to 2 decimal places
    aesthetics = c("color", "fill")  # Overrides dot outline and fill
  ) +
  theme_bw() + 
  theme(
    # Remove all grid lines
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
    
    # Remove title space
    plot.title = element_blank(),
    
    # Large text formatting
    axis.title = element_text(size = 18, face = "bold", color = "black"),
    axis.text.y = element_text(size = 16, color = "black"),
    axis.text.x = element_text(size = 16, color = "black"),
    
    # Large legend formatting
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.key.height = unit(1.2, "cm"),
    legend.key.width = unit(0.6, "cm")
  ) +
  guides(
    color = guide_colourbar(
      frame.colour = "black", 
      ticks.colour = "black", 
      order = 1
    ),
    fill = guide_colourbar(
      frame.colour = "black", 
      ticks.colour = "black", 
      order = 1
    )
  )


ggsave(
  filename = "C:\\Users\\amanhel\\OneDrive - The University of Liverpool\\Documents\\PhD\\papers\\dotplot_ora_geneRatio.svg", 
  plot = e_v2, 
  width = 12,       # Width in inches
  height = 8,       # Height in inches
  units = "in",
  device = "svg"
)

```
#Output
##Table of results
```{r}
head(go_enrich_reg_final_filt)
head(go_enrich_non_reg_final_filt)
```

## Upset Plot
Emphasizes the genes overlapping among different gene sets.
```{r fig.width=18, fig.height=12}
BiocManager::install("enrichplot")
library(enrichplot)
upsetplot(go_enrich_reg)
upsetplot(go_enrich_non_reg)
```

##Barplot

Similar to dotplot but shows gene count as bar length. Less information than dotplot since it doesn't show GeneRatio and p-value simultaneously. Redundant if you have the dotplot.

I will probably not use this, I have run it because it is in the tutorial, but the dotplot is better

```{r echo=TRUE}
barplot(go_enrich_reg, 
        drop = TRUE, 
        showCategory = 10, 
        title = "GO Biological Pathways",
        font.size = 8)
```

##Dotplot

Shows the top enriched GO terms with GeneRatio on x-axis, adjusted p-value as colour, and gene count as dot size

```{r echo=TRUE}
dotplot(go_enrich_reg, showCategory = 15, title = "Regulators - GO BP")
dotplot(go_enrich_non_reg, showCategory = 15, title = "Regulators - GO BP")
```

So ossification is way higher than cartilage development in the hierarchy (ossification is 3 and cartilage development is 16). When something is classified in one of teh child, it gets also classified as the parent so that is why they have a
very high percentage, and it makes sense as well

With simplify clusters these redundant terms together based on semantic similarity — how similar their gene annotations are — and keeps only the most representative term from each cluster, controlled by the cutoff parameter. A cutoff of 0.7 means terms that are 70% similar in their gene content get collapsed into one.

##Enriched GO induced graph:

Shows the GO hierarchy, how enriched terms relate to each other as a network or tre

To me this is interesting because you get why ossification has so much of the gene ratio
and count, it is a very high term so it makes sense it has a lot of them

```{r fig.width=12}
goplot(go_enrich_reg, showCategory = 10)
```



